
(4)

(b) Draw a sample of size $n = 400$ from $N(5, 1)$ in R. Compute $ν∗$ with the bootstrap replication
$B = 2000$ times. Plot the histogram of the bootstrap estimates $θ∗1, · · · , θ∗B$. This is a bootstrap
estimator for the distribution of $θ$.


```{r}

n <- 400
B <- 2000

xs <- rnorm(n, 5, 1)

sn <- sample(xs, n*B, replace = TRUE)

boot <- c()

for (i in 1:B) {
  boot[i] <- exp(mean(sn[(n*(i-1)+1):(n*i)]))
}

v_boot <- (1/B)*sum((boot - mean(boot))**2)

print(paste0("V_boot: ", round(v_boot, 2)))

```

```{r}
hist(boot, nclass=40)
```

(c) Draw 2000 samples (with fixed size $n = 400$) from $N(5, 1)$. For each sample, compute the estimate $θ_{bi}, i = 1, · · · , 2000$. Draw histogram of those $θ_{bi}$. This may be regarded as the true distribution of $θ_{b}$.
Compare it with its bootstrap estimator obtained in (b) above.

```{r}

n <- 400
B <- 2000

est <- c()
for (i in 1:B) {
  xs <- rnorm(n, 5, 1)
  sn <- sample(xs, n, replace = TRUE)
  est[i] <- exp(mean(sn))
}

v_boot_2 <- (1/B)*sum((est - mean(est))**2)

print(paste0("V_boot: ", round(v_boot_2, 2)))
```

```{r}
hist(est, nclass = 40)
```


(5)

(b)

```{r}
R <- 500
alpha <- 0.05

normal <- 0
pivot <- 0
percentile <- 0

for (j in 1:R) {

  n <- 100
  B <- 200
  
  xs <- rnorm(n, 1, 1)
  
  xs_hat <- mean(xs)/sd(xs)
  
  sn <- sample(xs, n*B, replace = TRUE)
  
  boot <- c()
  
  for (i in 1:B) {
    boot[i] <- mean(sn[(n*(i-1)+1):(n*i)])/sd(sn[(n*(i-1)+1):(n*i)])
  }
  
  v_boot <- (1/(B-1))*sum((boot - mean(boot))**2)
  
  l_boot <- sort(boot)[(alpha/2)*B]
    
  u_boot <- sort(boot)[(1 - alpha/2)*B]
  
  if ((xs_hat - sqrt(v_boot)*1.96 < 1) & (xs_hat + sqrt(v_boot)*1.96 > 1)) {
    normal <- normal + 1
  }
  
  if ((l_boot < 1) & (u_boot > 1)) {
    percentile <- percentile + 1
  }
  
  if ((2*xs_hat - u_boot < 1) & (2*xs_hat - l_boot > 1)) {
    pivot <- pivot + 1
  }

}

cat(paste0("Coverage\n",
             "Normal: ", (normal / 500)*100, "%\n",
             "Percentile: ", (percentile / 500)*100, "%\n",
             "Pivot: ", (pivot / 500)*100, "%\n"))

```


Revision

(4)

(b)
```{r}
n <- 400
B <- 2000

xs <- rnorm(n, 5, 1)

thetas <- c()
for (i in 1:B) {
  ys <- sample(xs, n, replace = TRUE)
  theta <- exp(mean(ys))
  thetas <- c(thetas, theta)
}

v_boot <- var(thetas)
hist(thetas)
```

(c)
```{r}
n <- 400
B <- 2000

thetas <- c()
for (i in 1:B) {
  ys <- rnorm(n, 5, 1)
  theta <- exp(mean(ys))
  thetas <- c(thetas, theta)
}

v <- var(thetas)
hist(thetas)
```


(5)

(b)
```{r}


normal <- 0
percentile <- 0
pivot <- 0

for (j in 1:500) {

n <- 100
B <- 200

xs <- rnorm(n, 1, 1)

rhos <- c()
for (i in 1:B) {
  ys <- sample(xs, size = n, replace = TRUE)
  rho <- mean(ys) / sd(ys)
  rhos <- c(rhos, rho)
}

u_boot <- mean(rhos)
v_boot <- var(rhos)

# Confidence intervals
normal_lb <- u_boot - 1.96*sqrt(v_boot)
normal_ub <- u_boot + 1.96*sqrt(v_boot)

percentile_lb <- sort(rhos)[B*0.025]
percentile_ub <- sort(rhos)[B*0.975]

pivot_lb <- 2*u_boot - sort(rhos)[B*0.975]
pivot_ub <- 2*u_boot - sort(rhos)[B*0.025]

if (1 > normal_lb & 1 < normal_ub) {
  normal <- normal + 1
}

if (1 > percentile_lb & 1 < percentile_ub) {
  percentile <- percentile + 1
}

if (1 > pivot_lb & 1 < pivot_ub) {
  pivot <- pivot + 1
}

}

cat("Normal CI:", normal / 500)
cat("Percentile CI:", percentile / 500)
cat("Pivot CI:", pivot / 500)

```











