---
title: "Ch1_Exercises"
output: html_document
---

```{r}

```


1. 


(a)

Plot of the marginal density of y: 
```{r}
y <- seq(-7,10,.02)
dens <- 0.5*dnorm(y,1,2) + 0.5*dnorm(y,2,2)
plot(y, dens)
```

(b)

```{r}
# Computing density of y at y=1:
cat("p(y=1):", 0.5*dnorm(1,1,2) + 0.5*dnorm(1,2,2),"\n")

# Computing density of y|theta at y=1:
cat("p(y=1}theta):", dnorm(1,1,2))
```

(c)

Looking at how the posterior probability that $\theta = 1$ changes with sigma. 
```{r}
sigmas <- seq(0.1,3,0.01)
prior <- 0.5
likelihood <- dnorm(1,1,sigmas**2)
normalized <- 0.5*dnorm(1,1,sigmas**2) + 0.5*dnorm(1,2,sigmas**2)
posterior_theta <- prior*likelihood/normalized

plot(sigmas, posterior_theta)
```


4. 

(b)

```{r}
# Using Normal approximation N(0,14^2) of outcome - prediction spread
cat("P(favorite wins|point spread = 8)", 1 - pnorm(-8.5,0,13.86),"\n",
    "P(beat the spread|point spread = 8)", 1 - pnorm(-0.5,0,13.86),"\n",
    "P(beat the spread|point spread = 8 & favorite wins)", 
        (1 - pnorm(-0.5,0,13.86))/(1 - pnorm(-8.5,0,13.86)))

```

Where the equation in the third calculation comes from (assuming continuous point differentials, which is not reflective of reality in this case because points are discrete):
$$Pr(BS|PS=8,FW) = \frac{Pr(BS,FW|PS=8)Pr(PS=8)}{Pr(FW|PS=8)Pr(PS=8)}$$


5.

(a)

Coarse prior knowledge and assumptions:

1) Only about 1/4 of all US congressional seats are actually competitive in any given year (3/4 of all seats have no chance of being tied). Of competitive districts, I will assume that the proportion of votes one candidate gets is distributed uniformly from 0.4 to 0.6. 
2) I do not have any information on how closely population size is tied to the closeness of races. I will assume independence.
3) Smaller congressional districts should be more common than larger districts. Therefore, using the information that districts range in population from 50,000 to 350,000, I will make the simplifying assumption that each congressional race has 100,000 voters. 
4) I will assume the results of races are independent (no specific information).

```{r}
# Monte Carlo simulation to determine a conditional probability of tying given this information

N <- 100000
voters <- 100000
races <- 435
lb_comp = 0.4
ub_comp = 0.6
p_comp = 0.25

ties <- 0

for (i in 1:N) {
  samp <- sample(seq.int(lb_comp*voters:ub_comp*voters),
                 size=races*p_comp, replace=TRUE)
  if ((voters / 2) %in% samp) {
    ties <- ties + 1
  }
}

cat("P(ties):", round(100*(ties/N), digits = 2), "%")

```

(b)

We assume (unrealisically, given rapid population growth) that the average number of votes was the same today at in 1900-1992. 

We also assume that the number of votes is distributed uniformly [-99,99] in these close elections and do not make use of the fact that 6 elections were within 10 votes. Then we calculate the following probabilities:


```{r}
cat("P(ties):", round(49/20593*100, 2), "%")
```


9.

(a)

```{r}
time_start <- 0
time_end <- 420

x <- cumsum(rexp(n=1000,rate = 1/10))
entry_times <- x[x <= 420]

docs <- c(0,0,0)

time <- 0
queue <- c()
waiting_times <- c()

len <- length(entry_times) + 1

for (i in seq.int(1:len)) {

  if (i < length(entry_times)+1) {
      time_elapsed <- entry_times[i] - time
  }
  else {
    time_elapsed <- 99999
    time <- 420
  }
  
  while (time_elapsed > 0) {
    wait <- pmin(pmax(min(docs),0), time_elapsed)
    time <- time + wait
    time_elapsed <- time_elapsed - wait
    
    docs <- pmax(docs - wait,0)
  
    if (docs[1] == 0) {
      docs[1] <- runif(1,5,20)
      waiting_times[queue[1]] <- pmax(time - entry_times[queue[1]], 0)
      queue <- queue[-1]
    }
    else if (docs[2] == 0) {
      docs[2] <- runif(1,5,20)
      waiting_times[queue[1]] <- pmax(time - entry_times[queue[1]], 0)
      queue <- queue[-1]
    }
    else if (docs[3] == 0) {
      docs[3] <- runif(1,5,20)
      waiting_times[queue[1]] <- pmax(time - entry_times[queue[1]], 0)
      queue <- queue[-1]
    }
    
    if (length(queue) == 0) {
       if (i < length(entry_times)+1) {
        time <- time + time_elapsed
        docs <- pmax(docs - time_elapsed,0)
        time_elapsed <- 0
       }
      else {
        break
      }
    }
  }
  
  if (i < length(entry_times) + 1) {
    queue[length(queue) + 1] <- i
  }
  else {
    time <- time + max(docs)
  }

}


```


(b)

```{r}
time_start <- 0
time_end <- 420

patients <- c()
num_waited <- c()
wait_time <- c()
time_closed <- c()

for (j in 1:100) {
x <- cumsum(rexp(n=1000,rate = 1/10))
entry_times <- x[x <= 420]

docs <- c(0,0,0)

time <- 0
queue <- c()
waiting_times <- c()

len <- length(entry_times) + 1

for (i in seq.int(1:len)) {

  if (i < length(entry_times)+1) {
      time_elapsed <- entry_times[i] - time
  }
  else {
    time_elapsed <- 99999
    time <- 420
  }
  
  while (time_elapsed > 0) {
    wait <- pmin(pmax(min(docs),0), time_elapsed)
    time <- time + wait
    time_elapsed <- time_elapsed - wait
    
    docs <- pmax(docs - wait,0)
  
    if (docs[1] == 0) {
      docs[1] <- runif(1,5,20)
      waiting_times[queue[1]] <- pmax(time - entry_times[queue[1]], 0)
      queue <- queue[-1]
    }
    else if (docs[2] == 0) {
      docs[2] <- runif(1,5,20)
      waiting_times[queue[1]] <- pmax(time - entry_times[queue[1]], 0)
      queue <- queue[-1]
    }
    else if (docs[3] == 0) {
      docs[3] <- runif(1,5,20)
      waiting_times[queue[1]] <- pmax(time - entry_times[queue[1]], 0)
      queue <- queue[-1]
    }
    
    if (length(queue) == 0) {
       if (i < length(entry_times)+1) {
        time <- time + time_elapsed
        docs <- pmax(docs - time_elapsed,0)
        time_elapsed <- 0
       }
      else {
        break
      }
    }
  }
  
  if (i < length(entry_times) + 1) {
    queue[length(queue) + 1] <- i
  }
  else {
    time <- time + max(docs)
  }

}

patients[j] <- length(entry_times)
num_waited[j] <- sum((waiting_times > 0))
wait_time[j] <- mean(waiting_times[waiting_times > 0])
time_closed[j] <- time - 420
}


cat("Number of patients:", median(patients), "\n",
    "Number of patients who waited:", median(num_waited), "\n",
    "Average wait time for patients who waited:", median(wait_time), "\n",
    "Minutes closed after 4pm:", median(time_closed), "\n")
```











